---
title: "Getting Started"
author: "R DelRossi"
date: "1/8/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(gt)
library(tidyverse)
```

## What is gt?

The `gt` package is a powerful way to create presentation-quality tables from your R data. Getting started is incredibly easy, but once you have an initial table, it's also easy to add sophisticated table features.

To get started, be sure you have installed the `gt` package and loaded it with `library(gt)`. In this document, I also loaded the `tidyverse` metapackage, which adds a variety of capabilities to our arsenal.

## Creating Tables with gt

A highly appealing characteristic of `gt` is its ability to make a simple table with minimal of effort.

```{r}
mtcars %>% head() %>% gt()
```

The result of the call to `gt()` is a `gt_tbl` object that can be saved to a variable and refined with additional calls to functions it the `gt` package. This ability to save the table in specific states can be especially useful when making decisions about the table's appearance that are based on conditional logic.

For instance, in this next example I save the initial table to the variable `tbl` and then set the table's title to a value that's based on some simple conditional logic.

```{r}
tbl <- mtcars %>% head() %>% gt()

if (ncol(mtcars) > 10) {
  tbl <- tbl %>% 
    gt::tab_header(title = "This table has more than 10 columns")
} else {
  tbl <- tbl %>% 
    gt::tab_header(title = "This table has 10 or fewer columns")
}

tbl
```
